---
title: "Introduction to rvtable"
author: "Matthew Leonawicz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette:
  fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Introduction to rvtable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- rvtable.md is generated from rvtable.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(comment="#>", message=FALSE, warning=FALSE, error=FALSE, eval=FALSE, tidy=TRUE) # for html_vignette only
```

The `rvtable` package provides a data frame-based object format and functionality for storing and manipulating
random variables. The emphasis is on continuous random variables though discrete random variables are also handled.
The `rvtable`-class object is a special type of data frame or data table that
contains a column defining values representing or sampled from a probability distribution and a column of respective probabilities.
Additional columns are ID columns and represent factors or categorical variables.
While the package makes generous use of `dplyr`, `rvtable`-class objects also share the class, `data.table`,
which is useful for efficiently working with large tables.

It is helpful to make the connection between operations performed by code and the probability concepts they parallel.
The entirety of an rvtable as described above represents a joint probability distribution of
the continuous random variable whose values and probabilities are represented by the values and probabilities columns
and any categrorical variables whose levels propogate the ID columns.

Filtering rows of an rvtable to specific combinations of ID variable levels yields a subset of the original table
that represents a conditional distribution of the continuous random variable.
Similarly, integrating out, marginalizing over, or collapsing the table over the unique levels of an ID variable
is analogous to calculating a marginal distribution of the continuous random variable.

This package offers a collection of functions that assist with manipulating tables of continuous random variables
in these and other ways. There is a function for generating rvtables from different kinds of input data.
An rvtable can be in probability distribution or sample form.
This means that an rvtable will contain `Val` and `Prob` columns (default names) when representing a distribution,
but only the `Val` column once sampled from. Form is tracked by object attributes.
Functions available in the package offer seamless transitions between forms, carrying out various operations
such as sampling from densities and marginalizing over levels of categorical variables in an rvtable.
Users can maintain control over the resolution of the data via function arguments that correspond to any
sampling or empirical density estimation steps that a function may perform.

This introduction covers the following topics:

- Create variuous rvtable objects.
- Sample from a distribution-type rvtable, yielding a new, sample-type rvtable.
- Compute marginal distributions on an rvtable.
- Compute an inverse empirical probability mass function from an rvtable.

## Creating rvtable objects

There are several ways to create an rvtable from the constructor function, `rvtable`.
Below are examples of continuous and discrete random variables stored in rvtables.
In the first example, the input, `x`, is simply a numeric vector representing a sample.
If `density.args` is not specified, the defaults are those used by `density`.
See the help documentation for `density` for details.
The second call below limits the sample from `n=512` to `n=50` and smooths the distribution
by setting `adjust=2`. We can also check to see if an object has the `rvtable` class.


```{r rvtable_objects1}
library(rvtable)
# basic samples from continuous and discrete RVs
x <- rnorm(100)
rvtable(x)
rvtable(x, density.args=list(n=50, adjust=2))
is.rvtable(rvtable(x))
```

This reveals that when working with "continuous" random variables, the distribution is
always discretized to some degree. When high fidelity is required, increase `n`.
This is increasingly important when working with a long data manipulation chain requiring 
many iterations of density estimation or sampling, such as marginalizing over multiple
categorical variables or if it is critical to maintain the shape of a multimodal distribution
with high precision.

The default is to assume `x` is continuous. In the next case, it is more ambiguous.
While all values in this example are clearly discrete integers, they are still treated as
samples from a continuous probability distribution and will be modeled internally by `density` as such
unless `discrete=TRUE`.

For discrete random variables, probabilities associated with values of `x` can be passed to `y`.
Alternatively, `x` can be passed without `y` for the same purpose if `x` has the attribute `probabilities`.
Note that in either case, it is no longer necessary to explicitly pass `discrete=TRUE`.
The discrete nature of `x` is inferred either from its attributes or from `y`.
When `x` is a numeric vector, `discrete=TRUE` is only required when probabilities are not passed by `y`
or by `x` attributes and `x` is a basic unweighted sample (but possibly with repeating values).
At any time use `attributes` to inspect the attributes of an rvtable.


```{r rvtable_objects2}
x <- sample(1:10, size=10000, replace=TRUE, prob=sqrt(10:1))
rvtable(x, discrete=TRUE) # required
x <- 1:5
probs <- c(0.1, 0.2, 0.3, 0.15, 0.25)
rvtable(x, y=probs) # discreteness inferred from y
attr(x, "probabilities") <- probs
rv <- rvtable(x) # discreteness inferred from attributes
rv
attributes(rv)
```

Also note above that probabilities need not sum to one in the discrete case.
They are relative weights and will be rescaled internally by `sample` rather than
passed to `density` as is the case with continous random variables.

Increasingly the complexity of the input to `rvtable`,
we arrive at the more common ways to constuct an rvtable: from and existing data frame or data table.
First is a data frame of values with equal probability.

Note that by leaving out `discrete=TRUE` in the call to `rvtable` below,
as with a numeric input, `x`, the data frame input is also assumed to represent a continuous
random variable sample. It will be estimated as if a tiny sample from a
uniform pdf rather than a fully represented uniform pmf.

```{r rvtable_objects3}
x <- data.frame(Val=1:10, Prob=0.1)
rv1 <- rvtable(x)
rv1
attributes(rv1)
rv2 <- rvtable(x, discrete=TRUE)
rv2
attributes(rv2)
```

Lastly, we load the `data.table` package so we can make a data table instead of a data frame.
It doesn't change anything in terms of rvtables since the package always applies the `data.table` class.
The key change here is the addition of an ID column, `id`, representing a categorical variable.
The column names referring to values and their associated probabilities can also be changed.

```{r rvtable_objects4}
library(data.table)
x <- data.table(id=rep(LETTERS[1:2], each=10), v1=rep(1:10, 2), p1=c(c(10:1)^2, sqrt(1:10)))
rv <- rvtable(x, Val="v1", Prob="p1")
rv
attributes(rv)
```

In concluding, here are three important properites of `rvtable` worth being clear on:

- An rvtable is not created from a data frame or data table that contains a values column
but no corresponding probabilities column, as can be done when x is a basic numeric samples vector.

Creating an rvtable from an existing data frame or data table requires a distribution-type
input. The user is required to provide a table with both columns.

- `rvtable` always creates an rvtable in distribution form, regardless of input.

- `rvtable` ignores grouping. 

If you pass a data frame or data table whose rows have been grouped by any categorical variables, 
for example using `dplyr::group_by`, this grouping information is ignored.
Instead, `rvtable` forcibly groups by all categorical variables present in the input, regardless.
It is reasonable to assume that any present categorical variables are already distinguishing
different sets of values and probabilities in a table and it is not intended for the
`rvtable` constructor to also apply any maginalizing operations.
`rvtable` always separates all apparent combinations of ID variables' levels.

## Sampling rvtables

Sampling on an rvtable requires a density-type rvtable.
This is the more common form. Sampling results in a new rvtable object that is of sampling-type, meaning its
`Val` column represents raw sample values and there is no corresponding `Prob` column of probabilities tied to values.
This is the less common rvtable type and tends to be explicitly created and used at the end of a chain of
data manipulation steps when it finally comes time to do something with the data such as plot it or pass samples to
an analysis.

Sampling is done using `sample_rvtable`. Below, small samples are drawn from rvtables containing
continous and discrete random variables, respectively.

```{r sample_rvtable1}
# continuous RV
x <- rvtable(rnorm(1000))
x
attributes(x)$tabletype
y <- sample_rvtable(x, n=10)
y
attributes(y)$tabletype

# discrete RV
x <- rvtable(sample(1:100, 50), discrete=TRUE)
y <- sample_rvtable(x, n=10)
sample_rvtable(y, n=8, resample=TRUE)
```

In the above example, the `tabletype` entry in the attributes of `y` is different from that in `x`.
After sampling from the density-type rvtable, the resultant rvtable is in sample form.
In the discrete example, `sample_rvtable` is called on an rvtable that is already in sample form.
In this case samples are taken directly from the rvtable `Val` column unless `resample=TRUE`, which
forces the additonal step of re-estimation of the pmf or pdf prior to resampling.
This is especially useful in the continuous case when the user requires more control over the 
sampling distribution via `density.args` prior to resampling.

Returning to the data table from eariler, sampling is performed by group.

```{r sample_rvtable2}
x <- data.table(id=rep(LETTERS[1:2], each=10), v1=rep(1:10, 2), p1=c(c(10:1)^2, sqrt(1:10)))
rv <- rvtable(x, Val="v1", Prob="p1")
sample_rvtable(rv, n=5)
```

Aside from `x`, other arguments to `sample_rvtable` include:

- `resample` as described.
- `n`, the sample size.

The sample size defaults to 10,000 because the typical context for the `rvtable` package is 
emprical estimation of continuous probability distributions
with high-fidelity/flexibility. While sampling comes after density estimation, it is also helpful to retain
larger samples if densities are subsequently re-estimated.

- `interp`, which defaults to `TRUE`, applies linear interpolation between samples prior to sampling.
- `n.interp`, the sample size resulting from `interp=TRUE`, defaults to 100,000.
- `decimals` is available for rounding samples.
- `density.args` takes a list of arguments to pass to `density` to override defaults, just as in `rvtable`.

Like the constructor, `rvtable`, `sample_rvtable` also forcibly applies its sampling to each identifiable group
and does not acknowledge any existing grouping information that has been added to a table.
For singling out or collapsing over ID variables, we turn to the next section.

## Computing marginal distributions on rvtables

There is not much to say about conditional distributions because this is relatively trivial.
A conditional distribution of the random variable described by `Val` and `Prob` given, say, an ID variable being
equal to one or some subset of some of its levels, is obtained by simply subsetting the rvtable to those rows.

On the other hand, marginalizing over levels of an ID variable
to yield a marginal distribution of the random variable decribed by `Val` and `Prob`, is more complex.
This is done with `marginalize`, which is a wrapper around `merge_rvtable`.
While the latter is exported by the `rvtable` package and can make sense to use directly,
it is typical to use `marginalize`, specifically.

Below, `marginalize` is used to collapse an rvtable by integrating out or marginalizing over
categrorical variables. The `margin` argument describes which variables to marginalize over.
In the first example, `marginalize` collapses the rvtable over the levels of `id1` and `id2`.

```{r marginalize1}
x <- data.table(
  id1=rep(LETTERS[1:5], each=4),
  id2=factor(c("low", "high")),
  id3=rep(1:2, each=2),
  Val=rep(1:10, each=20), Prob=rep(sqrt(1:10), each=20))
rv <- rvtable(x)
marginalize(rv, margin=c("id1", "id2"))
```

`get_levels` can be used to quickly check the unique levels of an ID variable.
Note that ID columns do not have to be of class `factor` to be treated as ID/categorical/factor variables, etc.
Character vectors suffice.

A reason this is helpful is because when marginalizing over an ID variable,
it is possible to pass a vector of weights or probabilities when integration should account for weighted samples.
For example, an rvtable may have equal numbers of rows describing the probability distribution of a continuous
random variable conditional on each of two ID variable levels. However, it may be known by the user that
each of these levels do not contribute equally to the marginal distribution.
Weights must be passed in the same orders as the levels returned by `get_levels`.

```{r marginalize2}
get_levels(rv, "id1")
marginalize(rv, "id1", weights=c(1, 1.5, 2, 4, 1))
```

When weights need to be specified, `marginalize` can only collapse over one ID variable at a time.
If multiple variables must be integrated out, this can usually be done with a single call as shown,
but when weights are supplied, then `marginalize` must be called successively.
The current implementation of `marginalize` does not allow `weights` to take a list of weights vectors 
for each of multiple ID variables.

## Computing an inverse pmf

An inverse pmf can be computed using `inverse_pmf`.
Continuing with the previous rvtable, we obtain the pdf of `id1` conditioned on a range of values
of the continuous random variable described by the original rvtable as well as the other present
factors.
The `val` column is now categorical and takes on the levels associated with `id1` in the input.

```{r inverse_pmf}
y1 <- inverse_pmf(rv, val.range=c(5, 8), "id1")
y1

library(dplyr)
x2 <- filter(rv, id2=="low" & id3==1) %>% select(-id2, -id3) %>% rvtable
y2 <- inverse_pmf(x2, c(5,8), "id1")
y2
```

The difference between the two calls to `inverse_pmf` above is that in the second call,
the rvtable has first been subset to one combination of levels of `id2` and `id3`.
In both cases, the pmf of `id1` is conditional on values of `id2`, `id3` and the continuous
rnadom variable represented by `Val` and `Prob` in the original rvtable;
it is merely not including all levels of `id2` and `id3` in the second example.

For more information about the `rvtable` package, see the [help documentation](http://leonawicz.github.io/rvtable/)
